\hypertarget{running-crexx-entails-compiling}{%
\section{Running cRexx entails
compiling}\label{running-crexx-entails-compiling}}

All Rexx scripts you run with cRexx are compiled by \texttt{rxc} into
Rexx assembler code (.rxas) and then assembled into an .rxbin file,
which contains the Rexx bytecode for execution by the Rexx Virtual
Machine \texttt{rxvm}. This rxvm executable takes care of linking
separately compiled modules together and executing them, so one function
can find another.

\hypertarget{a-first-program---hello-world}{%
\section{A first program - Hello
World}\label{a-first-program---hello-world}}

Let's say you have a Rexx exec you would like to run. To not have any
surprises, it is of the \emph{hello world} kind. We have a file called
hello.rexx, containing:

\begin{verbatim}
/* rexx */
options levelb
say 'hello cRexx world!'
\end{verbatim}

When cRexx level `C' (for `Classic') is available, the `options levelb'
(on line 2) statement can be left out; for the moment, level B is all we
have, and the compiler will refuse to compile without it.

With all our cRexx executables on the PATH, we only need to do:

\begin{verbatim}
rxc hello
rxas hello
rxvm hello
\end{verbatim}

to see `hello cRexx world!' on the console.

It might be a good idea to make a shell script to execute these three
programs in succession, and perhaps call it `crexx'. But take into
account that this really is a very simple case, in which no built-in
functions are called. You can look into the generated rexx assembler
(hello.rxas) file:

\begin{verbatim}
/*
 * cREXX COMPILER VERSION : cREXX F0043
 * SOURCE                 : hello
 * BUILT                  : 2022-12-03 22:27:52
 */

.srcfile="hello"
.globals=0

main() .locals=1 .expose=hello.main
   .meta "hello.main"="b" ".int" main() "" ""
   .src 3:1="say 'hello cRexx world!'"
   say "hello cRexx world!"
   .src 4:1="return 0"
   ret 0
\end{verbatim}

and you can see here that the compiler actually has generated a `say'
assembler instruction for the Rexx `say' instruction. (Assembler became
a whole lot easier with Rexx assembler.) But we did not yet call any
function.

\hypertarget{using-built-in-functions}{%
\section{Using built-in functions}\label{using-built-in-functions}}

Most Rexx programs use the extremely well designed built-in functions.
Now with these functions written in Rexx, and not hidden in the compiler
somewhere, we must tell it to import those from the library where we put
them earlier during the build process. Let's say we want to add a
display of the current weekday to our hello program. This will now be:

\begin{verbatim}
/* rexx */
options levelb
import rxfnsb
say 'hello cRexx world!'
say 'today is' date('w')
return 0
\end{verbatim}

Never mind the import statement, which you will not need when cRexx
`Classic' level C is available. But in level B, we need this, because we
need the flexibility it affords our plans for the future.

We must tell the compiler where to find the signature of the date()
function, so it can check if we call it in the correct way, with the
right parameters. This is done with the -i switch, which points to the
directory containing the library - which is called `library', by the
way.

\begin{verbatim}
rxc -i ~/crexx-build/lib/rxfns hello
\end{verbatim}

The assembler runs unchanged, because it trusts the compiler to have
checked if the called function really sits in that library, and has the
right parameters - the right code to call it has been generated.

\begin{verbatim}
rxas hello
\end{verbatim}

To run it, we can employ the `rxvme' executable - this one is extended
with linked-in versions of all the functions in the library:

\begin{verbatim}
rxvme hello
\end{verbatim}

which yields:

\begin{verbatim}
hello cRexx world!
today is Saturday
\end{verbatim}

\hypertarget{building-a-standalone-executable}{%
\section{Building a standalone
executable}\label{building-a-standalone-executable}}

To end this short tour of running cRexx, we are taking that last example
and will build a standalone executable out of it. If your neighbour,
family member or sports club runs the same OS as you do, they can now
run your compiled Rexx program from a USB stick, without ever installing
anything.

For this, we need the next set of commands, expressed as a Rexx exec:

\begin{verbatim}
/* rexx compile a rexx exec to a native executable */
/* Classic Rexx and NetRexx compatible             */
crexx_home='~/crexx-build'
if arg='' then do
  say 'exec name expected.'
  exit 99
end
parse arg execName'.'extension
if extension<>'' then say 'filename extension ignored.'
'rxc  -i' crexx_home'/lib/rxfns' execName
'rxas' execName
'rxcpack' execName crexx_home'/lib/rxfns/library'
'gcc -o' execName,
'-lrxvml -lmachine -lavl_tree -lplatform -lm -L',
crexx_home'/interpreter -L'crexx_home'/machine -L',
crexx_home'/avl_tree -L'crexx_home'/platform'  execName'.c'
\end{verbatim}

This exec is delivered in the source tree, bin directory. At the moment
it can be run by classic rexx interpreters and NetRexx, it will soon be
runnable by cRexx itself. The exec works by compiling the Rexx program
specified (again without the .rexx file extension) to an .rxbin Rexx
bytecode file, which is then serialized to a C source file, containing
the cRexx Virtual Machine and the library rxbin files, by the rxcpack
command. It is a rather peculiar looking C source, but nevertheless it
will compile to a working executable, which is done by the last step in
the exec, here using the gcc compiler. And you will be able to run it
without the overhead of checking, compiling, tokenizing to bytecode and
linking, so it will be quite fast:

\begin{verbatim}
hello cRexx world!
today is Saturday
./hello  0.00s user 0.00s system 61% cpu 0.009 total
\end{verbatim}
