% this is not genereated from Markdown so the edits can be here
\section{Register based Virtual Machine}
\crexx{} uses a register based virtual machine.
A \index{register-based virtual machine} uses registers, which are small, fast, on-chip memory locations, to store and manipulate data. In a register-based virtual machine, the instructions explicitly reference the registers by number, and the values are loaded into and out of the registers as needed. This approach can provide faster performance because accessing registers is typically faster than accessing memory. In the \emph{rxvm} virtual machine, the compiler of the underlying C implementation assigns the hardware registers.

In contrast, a \index{stack-based virtual machine} uses a stack to store and manipulate data. In a stack-based virtual machine, the operands are pushed onto the stack, and the operators pop them off the stack, perform the operation, and push the result back onto the stack. The stack-based approach is simpler than the register-based approach because there are no explicit references to registers in the instructions, and the operands are automatically managed by the stack. However, the stack-based approach can be slower than the register-based approach because accessing memory is typically slower than accessing registers.

Some other differences between a register-based virtual machine and a stack-based virtual machine include:
\begin{itemize}
\item Register-based virtual machines typically use more memory for the registers than stack-based virtual machines use for the stack.
\item Register-based virtual machines may require more complex instruction encoding and decoding logic than stack-based virtual machines, which can impact the size and complexity of the virtual machine implementation.
\item Stack-based virtual machines can be easier to implement and optimize for certain types of operations, such as function calls and loops, where the data can be easily pushed onto and popped off the stack.
\item Register-based virtual machines can provide more flexibility in terms of register allocation and optimization, which can be particularly important for high-performance applications.
\end{itemize}

The execution environment of a \crexx{} program is a
threaded\footnote{an alternative, non-threaded executable is available
  under the \emph{\index{rxbvm}} name} virtual
machine that is designed for optimal performance. This virtual
machine, implemented in the \emph{rxvm} executable, executes
machine instructions produced by the \emph{rxc} \crexx{}
compiler, or written by hand, assembled into an .rxbin binary file by
the \emph{rxas} assembler.

\input{Logical-REXX-VM.tex}

\section{\crexx{} Virtual Machine instructions}

% \section{Machine Interface}
This section describes the processor-specific information for
\index{RXVM processors}. The instruction set can be seen as the ISA
(instruction set architecture) for
an RXVM processor, of which the microcode is implemented in the C99 language.

% \subsection{Processor Architecture}
\index{processor architecture}
\index{instruction set}
Programs intended to execute directly on the processor use the
RXVM instruction set and the
instruction encoding and semantics of this architecture.

An application program can assume that all instructions defined by the
architecture and that are neither privileged nor optional, exist and work
as documented.

To be ABI (application binary interface) conforming, the processor must implement the instructions of
the architecture, perform the specified operations, and produce the
expected results.  The ABI neither places performance constraints on
systems nor specifies what instructions must be implemented in
hardware.  A software implementation of the architecture conforms to
the ABI; likewise, the architecture could be implemented in hardware,
e.g. an FPGA.

\subsection{The Register}
\index{register}
RXVM is a Register based virtual machine, as opposed to a Stack based
VM. The number of registers is only limited by memory and, for
practical purposes, can be considered unlimited. The address size of
the fields in a register is, for the virtual machine implementations,
implied by the address size that the host OS can handle. For hardware
the size is undefined and can follow the hardware address generation
capacity.

\begin{lstlisting}[label=crexxregister,caption={The
\crexx{} Register implementation in C}]
struct value {
    /* bit field to store value status - these are explicitly set */
    value_type status;

    /* Value */
    rxinteger int_value;
    double float_value;
    void *decimal_value; /* TODO */
    char *string_value;
    size_t string_length;
    size_t string_buffer_length;
    size_t string_pos;
#ifndef NUTF8
    size_t string_chars;
    size_t string_char_pos;
#endif
    void *object_value;
\end{lstlisting}

\subsubsection{Conversions of register values}
\index{conversions, data type in assembler}
\index{say instruction, rxas}
\index{int_value, register}
The status field determines for instructions that expect a data type,
which field is the field to act upon. Conversions are possible but
never implicit. So for example, when the register contains an int
value in field \code{int_value}, but it needs to be printed with the rxas
\keyword{say} instruction, the
\keyword{0\index{itos}} instruction takes care of the conversion and the population of the
memory area the \code{*string_value} points to. 

\begin{lstlisting}[label=crexxregister,caption={Values of
the Status field}]
typedef union {
    struct {
        unsigned int type_object : 1;
        unsigned int type_string : 1;
        unsigned int type_decimal : 1;
        unsigned int type_float : 1;
        unsigned int type_int : 1;
    };
    unsigned int all_type_flags;
} value_type;
\end{lstlisting}

