\hypertarget{rexx-assembler-specification}{%
\section{REXX Assembler
Specification}\label{rexx-assembler-specification}}

\hypertarget{overview}{%
\subsection{Overview}\label{overview}}

{[}Work in Progress{]}

\href{Logical-REXX-VM}{Additional details of the RXVM virtual machine}

\hypertarget{features-required-to-support-crexx-languages}{%
\subsection{Features Required to Support cREXX
Languages}\label{features-required-to-support-crexx-languages}}

The following table maps REXX features to RXAS capabilities to explain
the motivation for RXAS capabilities and approach for implementing REXX
features. Details of the REXX Features themselves are documented in the
REXX specifications, and more details of the RXAS capabilities follow.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3718}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
REXX~Features
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RXAS~Capabilities
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implementation~Approach
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Available
\end{minipage} \\
\midrule()
\endhead
Program Flow & Branches & Generate loops in RX & Yes \\
Simple Variables & Local Registers & & Yes \\
STEM Variables & See Classes & STEMs implemented as class & N/A \\
PROCEDURES & Procedures & & Yes \\
SIGNAL / Labels & branches / labels & may need duplicate code & Yes \\
Static EXPOSE & Pass by Reference args & EXPOSED vars by reference &
Yes \\
Dynamic EXPOSE & No Additional & Pool implemented in REXX & N/A \\
VALUE() & No Additional & Pool implemented in REXX & N/A \\
INTERPRET & Dynamically created rxbin & Complex\footnote{Version 1 will
  require the compiler / assembler linked into the runtime. Version 2
  (REXX on REXX) will require advanced REXX parsing.} & No \\
REXXb Classes & Regs contain sub-regs & \footnote{Class Attributes
  (always private) are stored in sub-registers. Member functions are
  statically linked and use a naming convention ``class.member''. The
  first argument to the member is the object} & No \\
REXXc Singleton Classes & global regs & \footnote{Singleton Classes
  replace static classes / members in other languages. The singleton is
  stored in a global register and exposed with a naming convention
  ``class.@1''. Note that ``@'' cannot be used from REXX programs
  directly} & Yes \\
REXXb Interface & Regs hold function pointer & No & \\
Dynamic / Late binding & footnote & by REXXb class(s) & No \\
Arbitrary Precision Maths & No Additional & in REXXb class(s) & N/A \\
ADDRESS & Platform specific & & No \\
External Functions & Platform specific & \footnote{Dynamic discovery and
  linking, call-backs to access variable pool. Classic REXX
  implementation will need to ensure all required variables are
  available in the variable pool} & No \\
Error Reporting & Annotation and debug pool & \footnote{Source line
  information stored in a ``debug'' pool to allow source line error
  reporting. Run time error conditions (signals) call REXX Exit
  functions} & No \\
\bottomrule()
\end{longtable}

\hypertarget{rxas-source-syntax-and-structure}{%
\subsection{RXAS Source Syntax and
Structure}\label{rxas-source-syntax-and-structure}}

\hypertarget{structure}{%
\subsubsection{Structure}\label{structure}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Global Variable definition or Declaration
\item
  Procedure Definition or Declaration (repeated)
\end{enumerate}

\hypertarget{comments}{%
\subsubsection{Comments}\label{comments}}

\begin{verbatim}
/* Block Comment */

* Line Comment
\end{verbatim}

\hypertarget{instructions}{%
\subsubsection{Instructions}\label{instructions}}

These have the format OP\_CODE {[}ARG1{[},ARG2{[},ARG3{]}{]}{]} where
each argument can be a

\begin{itemize}
\tightlist
\item
  Register - e.g.~r0\ldots1 (for locals), a0\ldots n (for arguments) or
  g0\ldots n (for globals)
\item
  String - e.g.~``hello''
\item
  Integer - e.g.~5
\item
  Float - e.g.~5.0
\item
  Function - e.g.~proc()
\item
  Identifiers / Label - e.g.~label1
\end{itemize}

\hypertarget{directives}{%
\subsubsection{Directives}\label{directives}}

RXAS supports the following directives.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1233}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8767}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Directive
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule()
\endhead
.globals = \{INT\} & Defines the number of globals defined for the
file \\
.locals = \{INT\} & Defines the number of local registers in a
procedure \\
.expose = \{ID\} & Defines the exposed index of a global register
\footnote{or Procedure. These are used for linking between files/modules} \\
\bottomrule()
\end{longtable}

\hypertarget{file-scope-global-registers}{%
\subsubsection{File Scope Global
Registers}\label{file-scope-global-registers}}

\begin{verbatim}
.globals={int}
\end{verbatim}

Defines \{int\} global variable g0 \ldots{} gn. These can be used within
any procedure in the file.

\hypertarget{global-registers}{%
\subsubsection{Global Registers}\label{global-registers}}

Any global register marked as exposed is available to any file which
also has the corresponding exposed index/name.

\hypertarget{file-1}{%
\paragraph{File 1}\label{file-1}}

\begin{verbatim}
.globals=2            * 2 Global Registers
g0 .expose=namespace.var_name   * 
\end{verbatim}

\hypertarget{file-2}{%
\paragraph{File 2}\label{file-2}}

\begin{verbatim}
.globals=3            * 3 Global Registers
g2 .expose=namespace.var_name   * 
\end{verbatim}

In this case file 1 g0 is mapped to file 2 g2 under the index/name of
``namespace.var\_name''.

\hypertarget{file-scope-procedure}{%
\subsubsection{File Scope Procedure}\label{file-scope-procedure}}

The locals define how many local registers, r0 to r(locals-1), are
needed by the procedure.

\begin{verbatim}
* The ".locals" shows the procedure is defined in here
file_scope_proc() .locals=3
...
ret
\end{verbatim}

\hypertarget{global-procedures}{%
\subsubsection{Global Procedures}\label{global-procedures}}

Global Procedures can be called between file/modules.

\hypertarget{file-1-1}{%
\paragraph{File 1}\label{file-1-1}}

\begin{verbatim}
* The ".locals" shows the procedure is defined in here
proc() .locals=3 .expose=namespace.proc
ret
\end{verbatim}

\hypertarget{file-2-1}{%
\paragraph{File 2}\label{file-2-1}}

\begin{verbatim}
* No ".locals" here! Showing that the procedure is only being declared 
rproc() .expose=namespace.proc

main() .locals=3 
call rproc()
ret
\end{verbatim}

In this case main() in File 2, calls rproc() which is globally provided
under the index/name of ``namespace.proc''. In File 1, proc() is exposed
under this name and hence called from File 1.

Note: that ``namespace'' hints at the use of namespaces as part of
exposed names; this facility is used by the compiler to define classes.

Also, as shown names can be mapped - they don't have to be the same in
the source and in the target.

\hypertarget{rxas-capabilities-alphabetical}{%
\subsection{RXAS Capabilities
(alphabetical)}\label{rxas-capabilities-alphabetical}}

\hypertarget{branching-and-labels}{%
\subsubsection{Branching and Labels}\label{branching-and-labels}}

Within a procedure labels can be defined as branch targets. Conditional
and unconditional branch instructions can target these labels. The
following example shows a loop structure.

\begin{verbatim}
   ...                   * Code before loop
l75:                     * Loop start label
   igt r0,r1,r4          * Does a integer compare of r1 and r4 - puts true (1) or false (0) into r0
   brt l37,r0            * Branch if true to l37 (i.e. branch out of the loop)
   ...                   * Instructions in the loop
   inc r1                * Increment r1 (the loop counter)
   br l75                * Unconditional Branch to the start of the loop
l37:                     * Loop end Label 
   ...                   * Instructions following the loop
\end{verbatim}

\hypertarget{code-annotation-and-debug-pool}{%
\subsubsection{Code Annotation and Debug
Pool}\label{code-annotation-and-debug-pool}}

NOT IMPLEMENTED

cREXX needs to support appropriate error messages (including source line
number / text), breakpoints, and tracing. RXAS directives (.file and
.line) allow the source file, source line to be defined. The .clause
directive allow clause boundaries to be defined.

\begin{verbatim}
.file = "testfile.rexx"   * Source file name 
proc()   .locals=3
   .regname r1,a          * Maps r1 to an id for debugging purposes 
   .regname r2,b
   .line 9 "a=5; b=6"     * The line number and source string
   .clause                * REXX clause boundary
   load r1,5
   .clause
   load r2,6
   .line 10 "say a+b"
   .clause
   iadd r0,r1,r2
   itos r0
   say r0
   .line 11 "return"
   .clause
   ret
\end{verbatim}

The directives are processed at ``build time'' and the debug constant
pool is created which allow assembler instructions to be mapped back to
source lines. In this way error messages can be generated as if the
source file was being interpreted ``classically'' but with no run-time
overhead.

In addition, tracing/debugging is implemented by the VM machine using
the clause boundaries stored in the debug pool. The VM can set a
breakpoint by replacing the instruction at the appropriate address with
a breakpoint instruction. When the breakpoint is reached it uses the
clause boundary information to determine where the next breakpoint
should be set. Tools can be made available to allow a REXX programmer to
set a breakpoint at a REXX source code line number.

The debug pool also contains the information to display the rexx
variable name stored in a register.

Note that accessing debug information is a significant overhead as the
debug pool will need repeated searching, and will only be used for
debugging/tracing (or creating an error message) where performance is
not critical. The reason this approach is used is that when there is no
debugging in action there is no runtime performance overhead at all
(obviously the size of the rxbin binary file is made larger with the
debug pool).

\hypertarget{constant-pool}{%
\subsubsection{Constant Pool}\label{constant-pool}}

Each File/Module has a constant pool that stores: * String Constants *
Procedure Details * PTable information (mapping class to interface
procedures for a class and objects)

\hypertarget{dynamic-access-to-registers-including-arguments}{%
\subsubsection{Dynamic Access to Registers (including
Arguments)}\label{dynamic-access-to-registers-including-arguments}}

Dynamic access to a register is enabled by additional members of the
link family of instructions these allow a register to point to the same
value as a dynmically number primary register.

\begin{verbatim}
alink secondary_reg, arg_reg_num * Links secondary_reg to the argument register with number stored in the int value of arg_reg_num 
glink secondary_reg, global_reg_num * Links secondary_reg to the global register with number stored in the int value of global_reg_num 
\end{verbatim}

\hypertarget{dynamic-type-instructions}{%
\subsubsection{Dynamic Type
Instructions}\label{dynamic-type-instructions}}

The compiler will be able to manage which registers have what values
most of the time but there will be certain dynamic situations where the
value type or status is not known. To handle this the compiler use the
registers type flag:

\begin{verbatim}
gettp - gets the register type flag (op1 = op2.typeflag)
settp - sets the register type flag (op1.typeflag = op2)
setortp - or the register type flag (op1.typeflag = op1.typeflag || op2)
brtpt - if op2.typeflag true then goto op1
brtpandt - if op2.typeflag && op3 true then goto op1
\end{verbatim}

The typeflag is a 64bit integer and its usage is defined by convention
only \protect\hyperlink{crexx-calling-convention}{see cREXX Calling
Convention}.

\hypertarget{dynamic-procedure-pointers}{%
\subsubsection{Dynamic Procedure
Pointers}\label{dynamic-procedure-pointers}}

This capability is to support interfaces. Where the compiler knows the
object's class it can link statically to the correct member by using the
procedure name (i.e.~class\_name.member\_name(), however when the object
is only known to implement an interface (i.e.~its class is not known)
then the VM needs to dynamically link interface members to the object's
class specific implementation.

Each register, that contains an object whose call implements an
interface, has a pointer to a entry in the constant pool. This entry
allows the the interface name and member number to be searched at
runtime, returning its implementation procedure pointer. This is known
as the register's static ptable. The static ptable also stores the name
of the objects class.

\emph{Note: Where an object members are dynamically assigned at runtime
(not an immediately required capability) the dynamic mapping from member
name to procedure pointer will be done within the REXX runtime library
(i.e.~not applicable to RXAS).}

A directive defines the entry

\begin{verbatim}
.ptable class_name interface1_name(impl1_1(), impl1_2(), ...) interface2_name(impl2_1() ...) ...
\end{verbatim}

This creates the entry into the constant pool with links to the
procedure implementing an interfaces members \#1,\#2 etc.

The object is linked to the entry with an instruction:

\begin{verbatim}
setptable r1,class_name
\end{verbatim}

This sets the register r1 ptable to the entry ``class\_name'' in the
constant pool.

Finally the entry can be used at runtime:

\begin{verbatim}
srcptable r2,"interface_name",3
\end{verbatim}

In this example r2 is a class instance (object) implementing interface
``interface\_name''. This instruction looks up the object's procedure
implementing member \#3 of interface ``interface\_name'' and sets the
procedure pointer of r2 to this. Then

\begin{verbatim}
dyncall r0, r2, r3
\end{verbatim}

calls the procedure in r2, with arguments from r3 and puts the result in
r0.

\hypertarget{external-functions}{%
\subsubsection{External Functions}\label{external-functions}}

\hypertarget{injecting-dynamic-code}{%
\subsubsection{Injecting Dynamic Code}\label{injecting-dynamic-code}}

\hypertarget{instructions-1}{%
\subsubsection{Instructions}\label{instructions-1}}

\hypertarget{type-coding}{%
\paragraph{Type coding}\label{type-coding}}

Instructions have prefix to determine type: s=string, i=integer,
f=float, o=object

\hypertarget{maths}{%
\paragraph{Maths}\label{maths}}

As an example, the add family will have * iadd reg,reg,reg\\
* fadd reg,reg,reg * etc.

Each function just uses the corresponding registers value (int, float,
etc).

\hypertarget{load}{%
\paragraph{Load}\label{load}}

\begin{itemize}
\tightlist
\item
  {[}s/i/f/d{]}load - lost loads the corresponding type value only
\item
  load (i.e.~with no prefix) copies all values and the type flag to the
  target register
\end{itemize}

\hypertarget{conversion}{%
\paragraph{Conversion}\label{conversion}}

Converting means setting a value for a type based on the value on
another type in the same register, e.g. * itos reg - sets the string
value to the string representation of the integer value of reg * ftos
reg * stof reg - This converts the string to a float, or triggers a
signal if it can't

Note: this replaces prime/master.

\hypertarget{say-address-etc.}{%
\paragraph{SAY / ADDRESS etc.}\label{say-address-etc.}}

Where an instruction needs a string it will only have a string
``version''. For clarity we will have

\begin{itemize}
\tightlist
\item
  say
\item
  address
\end{itemize}

but there will not be a isay etc. Instead the compiler might need to do
a ``itos'' first.

\hypertarget{procedures-and-arguments}{%
\subsubsection{Procedures and
Arguments}\label{procedures-and-arguments}}

A procedures registers are independent to the caller's registers. What
happens is that the VM maps its registers to the registers in the
caller.

Each time a procedure/function is called a new ``stack frame'' is
provided. This means that the called function has its own set of
registers.

The function header defines how many registers (called locals) the
function can access - for practical purposes we can consider that any
number of registers can be assigned to a function.

In addition, each file defines a number of global registers that can be
shared between procedures.

In a function with `a' arguments, `n' locals, and `m' globals: * R0
\ldots{} R(n-1) - are local registers to be used by the function * R(n)
\ldots{} R(n+m-1) - are the global registers, i.e.~g(0) \ldots{} g(m-1)
* R(n+m) - holds the number of arguments (a) * R(n+m+1) \ldots{}
R(n+m+a) holds the arguments, i.e.~a(1) \ldots{} a(a)

This ordering allows a dynamic numbers of arguments.

\hypertarget{crexx-calling-convention}{%
\paragraph{cREXX Calling Convention}\label{crexx-calling-convention}}

All arguments within RXAS are pass by reference, therefore arguments
needs to be copied to another register if pass by reference is not
wanted. This approach is a way to support moves rather than copies -
especially important to avoid slow object and string copies.

It is mandatory to use this calling convention between REXX and RXAS
procedures. Although not necessary, it is recommended to also use this
convention between RXAS procedures.

In this convention the caller is responsible for setting argument
registers' typeflag. This is used to indicate if an optional argument is
present, and if a pass-by-value string or object argument needs
preserving.

The callee (procedure) is responsible for applying default values for
optional arguments, and for ensuring that pass-by-value arguments are
kept constant (so they are not changed, effecting the caller logic) if
required. The callee uses the typeflag for this.

\textbf{Register Type Flag Byte Values}

The register typeflag is used to optimise function arguments.

\begin{itemize}
\item
  Bit 1 - REGTP\_VAL - ONLY used for optional arguments; setting (1)
  means the register has a specified value
\item
  Bit 2 - REGTP\_NOTSYM - ONLY used for ``pass be value'' and ONLY large
  (strings, objects) registers; setting (2) means that it is not a
  symbol so does not need copying as, even if it is changed, the caller
  will not use its original value. Note: Small registers (int, float)
  are always copied as this is faster than setting and checking this
  flag; the REGTP\_NOTSYM flag is not set or read for integers and
  floats.
\end{itemize}

The following examples demonstrate the calling convention.

\textbf{Basic Call by Reference}

REXX Program

Annotated Generated RXAS

\textbf{Optional Call by Reference}

REXX Program

Annotated Generated RXAS

\textbf{Call By Value Integer and Optimisations}

In this example, REGTP\_NOTSYM is not used as the parameter is an
integer.

REXX Program

Annotated Generated RXAS

\textbf{Call By Value Strings and Optimisations (optional arguments)}

In this example, REGTP\_NOTSYM is used as the parameter is an string in
optional arguments.

REXX Program

Annotated Generated RXAS

\hypertarget{arbitrary-number-of-arguments-with}{%
\paragraph{Arbitrary number of arguments with
\ldots{}}\label{arbitrary-number-of-arguments-with}}

\textbf{TO BE IMPLEMENTED} (requires arrays)

\hypertarget{expose}{%
\paragraph{EXPOSE}\label{expose}}

\textbf{TO BE IMPLEMENTED}

Syntax candy to provide familiar (but not the same) EXPOSE experience
for REXX programmers.

In this example:

\begin{verbatim}
exp = 100

say test( "hello")
exit

test: procedure = .string expose exp = .int
  arg message = .string
  return message || exp
\end{verbatim}

Is converted by the compiler to:

\begin{verbatim}
exp = 100

say test(exp, "hello")
exit

test: procedure = .string 
  arg expose exp = .int, message = .string
  return message || exp
\end{verbatim}

This is designed to provide a familiar (but not the same) experience for
REXX programmers

\hypertarget{procedure-lookup-tables}{%
\subsubsection{Procedure Lookup Tables}\label{procedure-lookup-tables}}

\hypertarget{registers}{%
\subsubsection{Registers}\label{registers}}

\hypertarget{register-data}{%
\paragraph{Register Data}\label{register-data}}

Each register holds 5 values - String, float, integer and object, and a
type flag which is used to indicate which values are valid. Note that
arbitrary position maths is handled as objects.

In most cases it is for the compiler to decide what values to use, and
how/if to set the type flag. Only a few dynamic scenarios will need some
extended functions (see following). The type values are 0=unset,
1=string, 2=float, 4=integer, 8=object, 16=interface. Each register can
have multiple types set as valid. The compiler sets the valid types
explicitly with instructions - this is not an automatic runtime
capability. At runtime the VM has no need nor the \emph{ability} to
validate data. Any caching has to be achieved by compiler logic.

An Object has an pointer to its static ptable entry in the constant pool
as well as an array of sub-registers. These sub-registers are the
private attributes of the object.

\hypertarget{register-initialisation}{%
\paragraph{Register Initialisation}\label{register-initialisation}}

All registers are initialized on entry to a procedure on the register
``stack''. The rationale is that all the memory can be malloced at once
which is faster/safer. The pointers to globals and arguments are also
setup.

In addition a shadow set of pointers to the procedure's registers are
setup. These are used by the unlink instruction, they holding the
base/initial register pointer, and unlink sets the register pointer to
the pointer held in the respective shadow value.

Registers hold references to their parent/owner for memory freeing
purposes. The owner can change, for example the owner of a returned
register is set to the caller. When an object, stack frame, global pool
is being deleted the registers are also freed/deleted if the registers
owner is the same as the container being deleted.

\hypertarget{register-re-mapping-facilities}{%
\paragraph{Register Re-Mapping
Facilities}\label{register-re-mapping-facilities}}

There are a few scenarios where the contents of a register is needed in
another register number: a call requires the arguments to be in
consecutive registers, object sub-registers need to be copied to
registers, or accessing an arbitrary argument registers (i.e.~when the
number of arguments is only known at runtime).

Copying the contents of registers to achieve this would be slow (and for
large strings or objects, very slow). Also it is inconsistent because an
integer copy and an object pointer copy (which is a copy by reference)
behave differently (integers become independent but a change to the
object changes it in each ``copy'').

We provide 2 facilities to allow very fast and safe register moves:

\begin{itemize}
\item
  SWAP. This swaps two registers. This is very fast as it requires only
  6 pointer copies (3 swapping the two register pointers and 3 swapping
  the two shadow pointers). It allows the programmer to swap two
  registers (arguments, globals, registers) to get the register into a
  convenient register number (perhaps for a call). Doing the swap again
  restores the register numbers.
\item
  LINK/UNLINK. The link instruction makes two register numbers point to
  the same register (one is primary, the other secondary). Unlink makes
  the \emph{secondary} register revert back to it original state by
  using the shadow register pointer. Each instruction only requires one
  pointer copy.
\end{itemize}

The behaviour should normally be quite simple - however swapping or
linking already linked or swapped registers may cause complex outcomes.
Developers should consider the above behaviour descriptions to untangle
this!

\hypertarget{runtime-error-conditions-and-exit-functions}{%
\subsubsection{Runtime Error Conditions and Exit
Functions}\label{runtime-error-conditions-and-exit-functions}}

\hypertarget{shell-instructions}{%
\subsubsection{Shell Instructions}\label{shell-instructions}}

\hypertarget{sub-registers}{%
\subsubsection{Sub-Registers}\label{sub-registers}}

\hypertarget{super-instructions}{%
\subsubsection{Super Instructions}\label{super-instructions}}

Once we see what the code generated by the compiler looks like we may
combine some of these instructions in to super instructions for
performance reasons.

\hypertarget{variable-pool-in-rexx}{%
\subsection{Variable Pool in REXX}\label{variable-pool-in-rexx}}

In the current code for the VM we have registers which point to a
variable structure (that can contain string, integers etc). These can be
considered to be ``anonymous variables'' -- the compiler will assign a
register to hold a variable at compile time.

In this way ``80\%'' of the needs for REXX programs will be handled --
but not all. Some aspects of REXX needs dynamic variable names --
e.g.~Some EXPOSE scenarios, INTERPRET, VALUE, and the REXXCOMM / SAA
API. When these are needed the compiler needs to work in what I am
calling ``Pedantic'' mode. In this mode variables are also given a
string index in a variable pool, this index can be searched for
dynamically at runtime.

Each stack frame will also include its own variable pool. This is a
name/value index (via a HASH or TREE) whereby a variable can be found
via the index string. \emph{This will be implemented in REXXb.}

\begin{itemize}
\tightlist
\item
  When a procedure exits and the stack frame is torn down, the
  corresponding variable pool and variables also need freeing.
\item
  To facilitate EXPOSE, a variable pool index can be linked to a parent
  pool variable.
\item
  A register can point to an anonymous variable (as implemented today in
  the code) or mapped to a variable in the variable pool.
\end{itemize}
