\chapter{Unicode}
One of the design goals of \crexx{} is to support Unicode on all
platforms that enable its use. 
Unicode is an important standard that helps ensure that different computer systems and software programs can handle and display text correctly, regardless of language, script, or writing system.

Here are some specific reasons why Unicode is important:
\begin{description}
\item[Universal character encoding] Unicode provides a single, universal character encoding that can represent all characters used in modern writing systems, including those used in different languages, scripts, and symbols. This means that Unicode enables computers and software programs to exchange text in any language or script.
\item[Multilingual support] Unicode supports the display of text in many different languages, including those that use non-Latin scripts, such as Arabic, Chinese, Cyrillic, Devanagari, Hebrew, and many others.
\item[Interoperability] Unicode enables interoperability between different computer systems and software programs, allowing users to share and exchange text without worrying about compatibility issues.
\item[Accessibility] Unicode makes it easier to develop software and tools that can support multiple languages and scripts, which is important for ensuring accessibility and inclusivity in digital communication.
\item[Standardization] Unicode is a widely accepted standard that is
  used by many software developers, hardware manufacturers, and
  technology companies. This standardization helps ensure consistency
  and compatibility across different platforms and devices.
\end{description}
Overall, Unicode plays a critical role in facilitating communication
and information sharing across different cultures and languages, and
is essential for creating a more connected and inclusive digital
world.
